---
title: Infectious Disease in NL
author: Jorik Bot
date: '2020-05-08'
slug: infectious-disease-in-nl
draft: true
categories:
  - r
  - ggplot2
tags:
  - r
  - ggplot2
subtitle: ''
summary: ''
authors: []
lastmod: '2020-05-08T17:29:56+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

```{r library loading}
library("tidyverse")
library("ggpubr")
library("ggsci")
```


```{r data import}
death_data <- read_csv2(file = "cbs_doden.csv")
disease_data <- read_csv2(file = "cbs_ziekten.csv")
vac_data <- read_csv2(file = "cbs_vaccinatie.csv")
```

## Overview

At the start of the 20th century about 3 in 1000 people died of some kind of infection every year. Which disease was reponsible for this?

```{r totaal}
death_data %>%
  ggplot(aes(perioden, totaal_infectie))+
    geom_point()+
    geom_line()+
    scale_x_continuous(breaks = c(1920, 1940, 1960, 1980, 2000, 2020))+
    theme_pubr()+
    labs(x = "Year",
         y = "Deaths per 100k inhabitants",
         caption = "Source: CBS Statline",
         title = "Deaths caused by infectious diseases")
```

## Tuberculosis

Not the flu, but tuberculoses caused most of these deaths. Surprisingly, the majority of the decrease in deaths occurred before the discovery of effective antibiotics against TBC. This was achieved thru increasing living standards, and isolation of the infected people in sanitoria (horrible place to be). Both world wars are clearly trendbreakers. A relapse in living standards immediatly saw rising mortality. 
```{r tb absolute}
death_data %>%
  pivot_longer(cols = c("totaal_infectie", "tuberculose", "griep"), names_to = "doodsoorzaak", values_to = "aantal") %>%
  filter(doodsoorzaak != "griep") %>%
    ggplot(aes(perioden, aantal, color = doodsoorzaak))+
      geom_point()+
      geom_line()+
      geom_vline(xintercept = 1946, linetype = "dashed")+
      annotate(geom = "text", x = 1947, y = 310, label = "Antibiotic era", hjust = "left")+
      #annotate(geom = "segment", x = 1947, y = 310, xend = 2020, yend = 310, arrow = arrow(length = unit(2, "mm"))) +
      scale_x_continuous(breaks = c(1920, 1940, 1960, 1980, 2000, 2020))+
      scale_y_continuous()+
      scale_color_jco(labels = c("Total deaths", "TBC deaths"))+
      theme_pubr()+
      theme(legend.title = element_blank())+
      labs(x = "Year",
           y = "Deaths per 100k caused by infectious diseases",
           caption = "Source: CBS Statline")
```


The amount of people dying from TBC decreased quickly in the first half of the 20th century. But without a cure TBC remained responsible for over 60% of all deaths from infectious diseases. Curing TBC finally became possible when the antibiotic streptomycin was discovered in 1944. In the following decades more antibiotics were discovered, which further improved outcomes. Nowadays only ~2 people per million die of TBC in the netherlands. 
```{r tb relative}
death_data %>%
  mutate(rel_tbc = tuberculose / totaal_infectie,
         rel_griep = griep / totaal_infectie) %>%
    ggplot(aes(perioden, rel_tbc))+
      geom_point()+
      geom_line()+
      geom_vline(xintercept = c(1946, 1970), linetype = "dashed")+
      annotate(geom = "text", x = 1947, y = 0.77, label = "streptomycin", hjust = "left")+
      annotate(geom = "text", x = 1971, y = 0.77, label = "rifampicin", hjust = "left")+
      scale_x_continuous(breaks = c(1920, 1940, 1960, 1980, 2000, 2020))+
      scale_y_continuous(breaks = c(0.0, 0.25, 0.5, 0.75), labels = scales::percent)+
      scale_color_jco()+
      theme_pubr()+
      labs(x = "Year",
           y = "TBC percent of total deaths by infections",
           caption = "Source: CBS Statline")
```

## What about the flu?

We go over 100%, so CBS data is flawed? Or flu is excluded from total for some reason?

```{r flu relative}
death_data %>%
  mutate(rel_tbc = tuberculose / totaal_infectie,
         rel_griep = griep / totaal_infectie) %>%
    ggplot(aes(perioden, rel_griep))+
      geom_point()+
      geom_line()+
      scale_x_continuous(breaks = c(1920, 1940, 1960, 1980, 2000, 2020))+
      scale_y_continuous(labels = scales::percent)+
      scale_color_jco()+
      theme_pubr()+
      labs(x = "Year",
           y = "Flu death percent of total deaths by infections",
           caption = "Source: CBS Statline")
```
